---
title: "Vignette"
output: 
  github_document: 
    toc: true
---

```{r, echo = FALSE}
# rmarkdown::render("vignette.Rmd", output_format = "github_document", output_file = "README.md")
```
In this vignette, we want to show how to access APIs to retrieve data. We use two NHL repositories as examples: [NHL records](https://gitlab.com/dword4/nhlapi/-/tree/master) and [NHL stats](https://gitlab.com/dword4/nhlapi/-/blob/master/stats-api.md).  

## Required Packages  

To be able to access data from APIs, you should install and load the `httr`, `jsonlite`, and `tidyverse` packages.  
    ```{r}
library(httr)
library(jsonlite)
library(tidyverse)
```

## Functions  

### NHL records API
```{r}
baseurl_records <- "https://records.nhl.com/site/api"
```
/franchise (Returns id, ﬁrstSeasonId and lastSeasonId and name of every team in the history
of the NHL)  

```{r}
getfran <- function(){
  fullurl <- paste0(baseurl_records, "/", "franchise")
  fran <- GET(fullurl) %>% content("text") %>% fromJSON(flatten = TRUE)
  return(fran)
}
getfran() %>% head()
```
/franchise-team-totals (Returns Total stats for every franchise (ex roadTies, roadWins, etc))  
```{r}
getFranTeamTot <- function() {
  fullurl <- paste0(baseurl_records, "/", "franchise-team-totals")
  franTeamTot <- GET(fullurl) %>% content("text") %>% fromJSON(flatten = TRUE)
  return(franTeamTot)
}
getFranTeamTot() %>% head()
```
/site/api/franchise-season-records?cayenneExp=franchiseId=ID (Drill-down into season
records for a speciﬁc franchise)  
```{r}
getFranSeaRec <- function(franID) {
  fullurl <- paste0(baseurl_records, "/", "franchise-season-records?cayenneExp=franchiseId=", franID)
  franSeaRec <- GET(fullurl) %>% content("text") %>% fromJSON(flatten = TRUE)
  return(franSeaRec)
}
getFranSeaRec("20") %>% head()
```
/franchise-goalie-records?cayenneExp=franchiseId=ID (Goalie records for the speciﬁed
franchise)  
```{r}
getFranGoaRec <- function(franID) {
  fullurl <- paste0(baseurl_records, "/", "franchise-goalie-records?cayenneExp=franchiseId=", franID)
  franGoaRec <- GET(fullurl) %>% content("text") %>% fromJSON(flatten = TRUE)
  return(franGoaRec)
}
getFranGoaRec("20") %>% head()
```
/franchise-skater-records?cayenneExp=franchiseId=ID (Skater records, same interaction as
goalie endpoint)  
```{r}
getFranSkaRec <- function(franID) {
  fullurl <- paste0(baseurl_records, "/", "franchise-skater-records?cayenneExp=franchiseId=", franID)
  franSkaRec <- GET(fullurl) %>% content("text") %>% fromJSON(flatten = TRUE)
  return(franSkaRec)
}
getFranSkaRec("20") %>% head()
```

### NHL stats API  


```{r}
baseurl_stats <- "https://statsapi.web.nhl.com/api/v1/teams"
getStats <- function(expand = "", teamID = "", stats = ""){
  if (expand != ""){
    fullurl <- paste0(baseurl_stats, "?expand=", expand)
  } else if (teamID != ""){
    fullurl <- paste0(baseurl_stats, "?teamID=", teamID)
  } else if (stats != ""){
    fullurl <- paste0(baseurl_stats, "?stats=", stats)
  }
  stats <- GET(fullurl) %>% content("text") %>% fromJSON(flatten = TRUE)
  return(stats)
}
getStats(expand = "person.names") %>% head()
getStats(stats = "statsSingleSeasonPlayoffs") %>% head()

```


## Exploratory Data Analysis  

```{r}
franTot <- getFranTeamTot()
franTot <- franTot$data %>% select(-c("id", "activeFranchise", "firstSeasonId", "gameTypeId", "lastSeasonId"))
franStats <- getStats(expand = "person.names") 
franStats <- franStats$teams %>% select(c("locationName", "firstYearOfPlay", "franchiseId", "venue.city", "venue.timeZone.id", "venue.timeZone.tz", "division.name", "conference.name"))
combined <- full_join(franTot, franStats, by = "franchiseId") 
head(combined)
ggplot(combined, aes(x = homeWins, y = homeLosses)) + geom_point(aes(color = division.name))
```

