---
title: "Vignette"
output: 
  github_document: 
    toc: true
---

```{r, echo = FALSE, eval = FALSE}
rmarkdown::render("vignette.Rmd", output_format = "github_document", output_file = "README.md")
```
This vignette demonstrates how to access APIs to retrieve data. We use two NHL repositories as examples: [NHL records](https://gitlab.com/dword4/nhlapi/-/tree/master) and [NHL stats](https://gitlab.com/dword4/nhlapi/-/blob/master/stats-api.md).  

## Required Packages  

To be able to access data from APIs, you should install and load the `httr`, `jsonlite`, and `tidyverse` packages.  
```{r}
library(httr)
library(jsonlite)
library(tidyverse)
```

## Functions  

### NHL records API
```{r, message = FALSE}
baseurl_records <- "https://records.nhl.com/site/api"
```
The function `getFran` retrieves basic information (id, ï¬rstSeasonId and lastSeasonId and name of every team in the history of the NHL) about all teams. No arguments are needed.  

```{r, message = FALSE}
getFran <- function(){
  fullurl <- paste0(baseurl_records, "/", "franchise")
  fran <- GET(fullurl) %>% content("text") %>% fromJSON(flatten = TRUE)
  return(fran$data)
}
getFran() %>% tbl_df()

```
The function `getFranTeamTot` retrieves stats about all teams (Total stats for every franchise (ex roadTies, roadWins, etc)). No arguments are needed.  

```{r, message = FALSE}
getFranTeamTot <- function() {
  fullurl <- paste0(baseurl_records, "/", "franchise-team-totals")
  franTeamTot <- GET(fullurl) %>% content("text") %>% fromJSON(flatten = TRUE)
  return(franTeamTot$data)
}
getFranTeamTot() %>% tbl_df()
```

To allow for convenient access of team information in the following functions, we first construct a subset of data, so users can use team names or franchise ID (or team names or team ID) to look up information.  

```{r, message = FALSE}
index <- getFranTeamTot() %>% select(c("franchiseId", "teamName", "teamId")) %>% unique()
```

The function `getFranSeaRec` retrieves season records for a specific franchise, and therefore the argument `franchiseId` or `teamName` is needed. (`franchiseId` can be found using `getFran` or `getFranTeamTot`.)  

```{r, message = FALSE}
getFranSeaRec <- function(team) {
  if (is.character(team)) {
    id <- index$franchiseId[index$teamName == team]
  } else if (is.numeric(team)){
    id <- team
  }
  fullurl <- paste0(baseurl_records, "/", "franchise-season-records?cayenneExp=franchiseId=", id)
  franSeaRec <- GET(fullurl) %>% content("text") %>% fromJSON(flatten = TRUE)
  return(franSeaRec$data)
}
getFranSeaRec(20) %>% tbl_df()
getFranSeaRec("Vancouver Canucks") %>% tbl_df()
```
The function `getFranGoaRec` retrieves goalie records for a specific franchise, and again a `franchiseId` or `teamName` is required.  

```{r, message = FALSE}
getFranGoaRec <- function(team) {
  if (is.character(team)) {
    id <- index$franchiseId[index$teamName == team]
  } else if (is.numeric(team)){
    id <- team
  }
  fullurl <- paste0(baseurl_records, "/", "franchise-goalie-records?cayenneExp=franchiseId=", id)
  franGoaRec <- GET(fullurl) %>% content("text") %>% fromJSON(flatten = TRUE)
  return(franGoaRec$data)
}
getFranGoaRec(20) %>% tbl_df()
getFranGoaRec("Vancouver Canucks") %>% tbl_df()
```

The function `getFranSkaRec` retrieves information about skater records for a specific franchise, and an argument about `franchiseId` or `teamName` is required.  

```{r, message = FALSE}
getFranSkaRec <- function(team) {
  if (is.character(team)) {
    id <- index$franchiseId[index$teamName == team]
  } else if (is.numeric(team)){
    id <- team
  }
  fullurl <- paste0(baseurl_records, "/", "franchise-skater-records?cayenneExp=franchiseId=", id)
  franSkaRec <- GET(fullurl) %>% content("text") %>% fromJSON(flatten = TRUE)
  return(franSkaRec$data)
}
getFranSkaRec(20) %>% tbl_df()
getFranSkaRec("Vancouver Canucks") %>% tbl_df()
```

### NHL stats API  
  
For this function, eight modifiers can be chosen, and thus one or more arguments (`ID`, `expand`, `teamID`, `stats`, and/or `season`) has to be provided. Below are the eight modifiers:  

  * ?expand=team.roster Shows roster of active players for the specified team  
  * ?expand=person.names Same as above, but gives less info.  
  * ?expand=team.schedule.next Returns details of the upcoming game for a team  
  * ?expand=team.schedule.previous Same as above but for the last game played  
  * ?expand=team.stats Returns the teams stats for the season  
  * ?expand=team.roster&season=20142015 Adding the season identifier shows the roster for that season  
  * ?teamId=4,5,29 Can string team id together to get multiple teams  
  * ?stats=statsSingleSeasonPlayoffs Specify which stats to get. Not fully sure all of the values  

Examples of arguments:  

  * `ID = 20`
  * `ID = "Calgary Flames"`  
  * `expand = "person.names"`  
  * `teamId = "4, 5, 29"`  
  * `stats = "statsSingleSeasonPlayoffs"`  
  * 'expand = "team.stats", season = "20142015"`  

Note: If you would like information for more than one team, enter `teamId`, but, in this case, no more specific information can be chosen. If you want specific information, such as roster, enter only one `ID` or `teamName`.  

For more information about these modifiers, see [the documentation](https://gitlab.com/dword4/nhlapi/-/blob/master/stats-api.md).

```{r, message = FALSE}
baseurl_stats <- "https://statsapi.web.nhl.com/api/v1/teams"
getStats <- function(ID = "", expand = "", teamID = "", stats = "", season = ""){
  if (ID != ""){
      if (is.character(ID)) {
        ID <- index$teamId[index$teamName == ID]
      }
    baseurl_stats <- paste0("https://statsapi.web.nhl.com/api/v1/teams/", ID)
  }
  if (teamID != ""){
    fullurl <- paste0("https://statsapi.web.nhl.com/api/v1/teams?teamId=", teamID)
  } else if (expand != ""){
    if (expand == "team.roster" || expand == "team.stats"){
       if (season != ""){
         fullurl <- paste0(baseurl_stats, "?expand=", expand, "&season=", season)
       } 
    } else {
    fullurl <- paste0(baseurl_stats, "?expand=", expand)
    } 
  } else if (stats != ""){
    fullurl <- paste0(baseurl_stats, "?stats=", stats)
  }
  stats <- GET(fullurl) %>% content("text") %>% fromJSON(flatten = TRUE)
  if (str_detect(expand, "team.roster")){
    stats_new <- stats$teams$roster.roster[[1]]
    stats <- cbind(stats$teams[,c(1,2,10)], stats_new)
  } else if (expand == "team.schedule.next"){
    if (is.null(stats$teams$nextGameSchedule.dates)){
      stop("No information is available")
    } else {
    stats_new <- stats$teams$nextGameSchedule.dates
    stats <- cbind(stats$teams[1,c(1,2,10)], stats_new)
    }
  } else if (expand == "team.schedule.previous"){
    if (is.null(stats$teams$previousGameSchedule.dates)){
      stop("No information is available")
    } else {
    stats_new <- stats$teams$previousGameSchedule.dates
    stats <- cbind(stats$teams[,c(1,2,10)], stats_new)
    }
  } else if (expand == "team.stats"){
    stats_new <- stats$teams$teamStats[[1]]$splits[[1]]
    stats <- cbind(stats$teams[1,c(1,2,11)], stats_new)
  } else {
    stats <- stats$teams
  }
  if (is.null(stats)) {
    stop("No information is available")
  } 
  return(as.data.frame(stats))
}
# getStats(ID = 20, expand = "team.stats")
# getStats(ID = "Calgary Flames", expand = "person.names")
getStats(ID = 14, expand = "team.schedule.next") %>% tbl_df()
getStats(ID = 20, expand = "team.stats", season= "20102011") %>% tbl_df()
# getStats(ID = 53, expand = "team.roster") 
getStats(teamID = "4,5,29") %>% tbl_df()
# getStats(ID = 54, stats = "statsSingleSeasonPlayoffs") %>% tbl_df()

```
### A wrapper function for all the functions above  

Endpoints:  

  * franchise (getFran)  
  * team total (getFranTeamTot)  
  * season record (getFranSeaRec)  
  * goalie record (getFranGoaRec)  
  * skater record (getFranSkaRec)  
  * stats (getStats)  
    + expand = "team.roster"  
    + expand = "person.names"  
    + expand = "team.schedule.next"  
    + expand = "team.schedule.previous"  
    + expand = "team.stats"  
    + expand = "team.roster", season = "20142015"  
    + teamId = "4, 5, 29"  
    + stats = "statsSingleSeasonPlayoffs"  

See more details in the documentation of [NHL records](https://gitlab.com/dword4/nhlapi/-/blob/master/records-api.md) and [NHL stats](https://gitlab.com/dword4/nhlapi/-/blob/master/stats-api.md).

The function `nhlFun` can retrieve any of the datapoints covered above, and one or more arguments listed above are required.  

Examples of arguments:  

  * `endpoint = "stats", ID = "Calgary Flames", expand = "team.roster"`  
  * `endpoint = "franchise"`  

```{r, message = FALSE}
# add a 
nhlFun <- function(endpoint, ...){
  if (endpoint == "franchise") {
    getFran()
  } else if (endpoint == "team total") {
    getFranTeamTot()
  } else if (endpoint == "season record"){
    getFranSeaRec(...)
  } else if (endpoint == "goalie record"){
    getFranGoaRec(...)
  } else if (endpoint == "skater record"){
    getFranSkaRec(...)
  } else if (endpoint == "stats"){
    getStats(...)
  } else {
    stop("Please enter a valid endpoint")
  }
}
# nhlFun(endpoint = "skater record", 20) %>% tbl_df()
nhlFun(endpoint = "stats", ID = "Calgary Flames", expand = "person.team") %>% tbl_df()
nhlFun(endpoint = "stats", ID = 14, expand = "team.schedule.next") %>% tbl_df()
```

## Exploratory Data Analysis  

### Retrieve Information

Now we use functions above to do exploratory data analysis. We will first retrieve data from two endpoints: **team total** and **person.names**, remove some columns we don't need, and combine the two data frames. Then we'll add two new variables: `winPercent` (the proportion of wins among all games played) and `homeWinPercent` (the proportion of wins of home games among all home games). We also create a subset to include fewer columns and transform the data from wide to long to show the numbers of wins, losses, and ties in various situations for each division.  

```{r, message = FALSE}
# perhaps use the wrapper function here
franTot <- nhlFun(endpoint = "team total")
franTot <- franTot %>% select(-c("id", "activeFranchise", "firstSeasonId", "lastSeasonId"))
franStats <- nhlFun(endpoint = "stats", expand = "person.names") 
franStats <- franStats %>% select(c("locationName", "firstYearOfPlay", "franchiseId", "venue.city", "venue.timeZone.id", "venue.timeZone.tz", "division.name", "conference.name"))
# create two variables: winPercent and homeWinPercent 
combined <- inner_join(franTot, franStats, by = "franchiseId") %>% mutate(winPercent = wins / gamesPlayed, homeWinPercent = homeWins / (homeWins + homeLosses + homeOvertimeLosses + homeTies))
head(combined)

# create a subset for numbers of games lost or won
subset <- combined %>% select(starts_with("home"), starts_with("road"), "division.name", -"homeWinPercent") %>% gather(-"division.name", key = "type", value = "game") %>% group_by(division.name, type) %>% summarise(sum = sum(game, na.rm = TRUE))
```

### Summaries  

Now we have the data, we can see some summary statistics for continuous variables. It is interesting to see that, for all the variables listed, the medians are much smaller than the means, suggesting that the distributions are right skewed. Perhaps a few teams have larger than average values, pulling the means upwards. A plot on penalty minutes shows the right-skewedness.  

Two contingency tables show the counts of two categorical variables. The first contingency table shows that 1926, 1967, and 1979 added the most number of teams. The second contingency table show that the same numbers of teams played in the regular seasons and playoffs in the Atlantic, Central, and Pacific seasons. Somehow one team in the Metropolitan division did not show up in the playoffs. Perhaps the data is missing or the team really did not make it into the playoffs.  

```{r, message = FALSE}
# summaries
apply(combined[,c(3, 6:12, 14:18)], FUN = summary, MARGIN = 2)
# contingency table
table(combined$division.name, combined$firstYearOfPlay)
table(combined$division.name, combined$gameTypeId)
```


### Visualize Data

Now we have the data and summaries, we can make plots to visualize the data.  

In the first scatter plot, we can see that as a lot of teams played low numbers of games, and a few teams played higher numbers of games. The winning percentage increases slightly as the number of games played increases, but the slope is small and the difference might not be significant. The colors of dots represent different divisions, and it does not appear that there are different patterns for different divisions. The second plot looks at each division separately; again the patterns for different divisions do not seem to differ. The second plot also shows that no teams in the Pacific division played more than 4000 games, but a few teams in all the other divisions played more than 4000 games.  

The third plot is similar, but the y-axis shows the winning percentage of home games, and the colors represent different types of games (regular season vs. playoffs). It does not appear that the number of games played and the winning percentage of home games are correlated, as the slope is close to zero.  

```{r, message = FALSE}
# scatter plot of homeWins and roadWins
ggplot(combined, aes(x = gamesPlayed, y = winPercent)) + geom_point(aes(color = division.name), position = "jitter") + geom_smooth(method = lm, color = "blue") + scale_color_discrete(name = "Division Name") + ggtitle("Scatterplot: games played vs. winning percentages of all games") + xlab("Games Played") + ylab("Winning percentages of all games")

ggplot(combined, aes(x = gamesPlayed, y = winPercent)) + geom_point(position = "jitter") + geom_smooth(method = lm, color = "blue") + facet_wrap(~ division.name) + ggtitle("Scatterplot: game played vs. winning percentages of all games by division") + xlab("Games Played") + ylab("Winning percentages of all games")

ggplot(combined, aes(x = gamesPlayed, y = homeWinPercent)) + geom_point(aes(color = as.factor(gameTypeId)), position = "jitter") + geom_smooth(method = lm, color = "blue") + scale_color_discrete(name = "Game Type", labels = c("regular season", "playoffs")) + ggtitle("Scatterplot: games played vs. winning percentages of home games") + xlab("Games Played") + ylab("Winning percentages of home games")
```

The histogram shows that the center of winning percentages of all teams is around 0.5, and the distribution is left-skewed. Therefore most of the teams have winning percentages between 0.3 and 0.6, but some teams have much lower winning percentages.  

```{r}
# histogram of winPercent
ggplot(combined, aes(x = winPercent)) + geom_histogram(bins = 30, aes(y = ..density..)) + geom_density(kernel = "gaussian", color = "red", alpha = 0.5, fill = "green") + ggtitle("Histogram of winning percentages of all games") + xlab("Winning percentages of all games")
```

Next we look at the total numbers of games played for each division. The boxplots show that, on average, the Central division played the most games, and the Pacific division played the fewest. However, the variation is large, and therefore the differences might not be significant. Moreover, the patterns of venues of different time zones do not seem to differ.  

```{r, message = FALSE}
# boxplots of gamesPlayed by division
ggplot(combined, aes(x = division.name, y = gamesPlayed)) + geom_boxplot() + geom_jitter(aes(color = venue.timeZone.tz)) + ggtitle("Boxplots of games played by division") + xlab("Division name") + ylab("Games played") + scale_color_discrete(name = "Venue time zone")
```

Now we look at two barplots. The first one shows wins, losses, and ties of each division. We see that the Atlantic division play the highest number of games and the Pacific division play the lowest number of games. Overall, most results are from wins and losses, and ties and overtime occupy a small proportion for all divisions. The second barplots show that all divisions have higher winning percentages in playoffs than in the regular season. The differences among divisions might not be significant.  

```{r, message = FALSE}
# barplot of gamePlayed
ggplot(subset, aes(y = sum, fill = type)) + geom_bar(position = "stack", stat = "identity", aes(x = division.name)) + ggtitle("Barplot of wins, losses, and ties by division") + xlab("Division name") + ylab("Total games played") + scale_fill_discrete(name = "Game Type")

ggplot(combined, aes(x = division.name, y = winPercent, fill = as.factor(gameTypeId))) + geom_bar(position = "dodge", stat = "identity") + scale_fill_discrete(name = "Game Type", labels = c("regular season", "playoffs")) + ggtitle("Barplot of winning percentages by division") + xlab("Division name") + ylab("Winning percentages of all games")
```

Lastly, we look at a frequency plot of penalty minutes. The plot clearly shows that the distribution is right-skewed. 20 teams had no penalty minutes, and a few teams had penalty minutes of more than 50,000.  

```{r, message = FALSE}
# frequency plot for penalty minutes
ggplot(combined, aes(x = penaltyMinutes)) + geom_freqpoly() + ggtitle("Frequency plot of penalty minutes") + xlab("Penalty minutes")
```

