---
title: "Vignette"
output: 
  github_document: 
    toc: true
---

```{r, echo = FALSE, eval = FALSE}
rmarkdown::render("vignette.Rmd", output_format = "github_document", output_file = "README.md")
```
This vignette demonstrates how to access APIs to retrieve data. We use two NHL repositories as examples: [NHL records](https://gitlab.com/dword4/nhlapi/-/tree/master) and [NHL stats](https://gitlab.com/dword4/nhlapi/-/blob/master/stats-api.md).  

## Required Packages  

To be able to access data from APIs, you should install and load the `httr`, `jsonlite`, and `tidyverse` packages.  
```{r}
library(httr)
library(jsonlite)
library(tidyverse)
```

## Functions  

### NHL records API
```{r, message = FALSE}
baseurl_records <- "https://records.nhl.com/site/api"
```
`getFran` retrieves basic information about all teams.  

/franchise (Returns id, ï¬rstSeasonId and lastSeasonId and name of every team in the history
of the NHL)  

```{r, message = FALSE}
getFran <- function(){
  fullurl <- paste0(baseurl_records, "/", "franchise")
  fran <- GET(fullurl) %>% content("text") %>% fromJSON(flatten = TRUE)
  return(fran$data)
}
getFran() %>% tbl_df()

```
`getFranTeamTot` retrieves stats about all teams.  

/franchise-team-totals (Returns Total stats for every franchise (ex roadTies, roadWins, etc))  

```{r, message = FALSE}
getFranTeamTot <- function() {
  fullurl <- paste0(baseurl_records, "/", "franchise-team-totals")
  franTeamTot <- GET(fullurl) %>% content("text") %>% fromJSON(flatten = TRUE)
  return(franTeamTot$data)
}
getFranTeamTot() %>% tbl_df()
```
To allow for convenient access of team information in the following functions, we first construct a subset of data, so users can use team names or franchise ID to look up information.  

```{r}
index <- getFranTeamTot() %>% select(c("franchiseId", "teamName", "teamId")) %>% unique()
```

`getFranSeaRec` retrieves season records from one specific team, and therefore you need to provide the `franchiseId` or `teamName` as an argument. `franchiseId` can be found using `getFran` or `getFranTeamTot`.  

/site/api/franchise-season-records?cayenneExp=franchiseId=ID (Drill-down into season
records for a specific franchise)  

```{r, message = FALSE}
getFranSeaRec <- function(team) {
  if (is.character(team)) {
    id <- index$franchiseId[index$teamName == team]
  } else if (is.numeric(team)){
    id <- team
  }
  fullurl <- paste0(baseurl_records, "/", "franchise-season-records?cayenneExp=franchiseId=", id)
  franSeaRec <- GET(fullurl) %>% content("text") %>% fromJSON(flatten = TRUE)
  return(franSeaRec$data)
}
getFranSeaRec(20) %>% tbl_df()
getFranSeaRec("Vancouver Canucks") %>% tbl_df()
```
`getFranGoaRec` retrieves goalie records, and again a `franchiseId` or `teamName` is required.  

/franchise-goalie-records?cayenneExp=franchiseId=ID (Goalie records for the specified
franchise)  

```{r, message = FALSE}
getFranGoaRec <- function(team) {
  if (is.character(team)) {
    id <- index$franchiseId[index$teamName == team]
  } else if (is.numeric(team)){
    id <- team
  }
  fullurl <- paste0(baseurl_records, "/", "franchise-goalie-records?cayenneExp=franchiseId=", id)
  franGoaRec <- GET(fullurl) %>% content("text") %>% fromJSON(flatten = TRUE)
  return(franGoaRec$data)
}
getFranGoaRec(20) %>% tbl_df()
getFranGoaRec("Vancouver Canucks") %>% tbl_df()
```

`getFranSkaRec` retrieves information about skater records, and a `franchiseId` or `teamName` is required.  

/franchise-skater-records?cayenneExp=franchiseId=ID (Skater records, same interaction as
goalie endpoint)  

```{r, message = FALSE}
getFranSkaRec <- function(team) {
  if (is.character(team)) {
    id <- index$franchiseId[index$teamName == team]
  } else if (is.numeric(team)){
    id <- team
  }
  fullurl <- paste0(baseurl_records, "/", "franchise-skater-records?cayenneExp=franchiseId=", id)
  franSkaRec <- GET(fullurl) %>% content("text") %>% fromJSON(flatten = TRUE)
  return(franSkaRec$data)
}
getFranSkaRec(20) %>% tbl_df()
getFranSkaRec("Vancouver Canucks") %>% tbl_df()
```

### NHL stats API  
  
For this function, eight modifiers can be chosen, and thus an argument (`expand`, `teamID`, or `stats`) has to be provided. Below are the eight modifiers:  

  * ?expand=team.roster Shows roster of active players for the specified team  
  * ?expand=person.names Same as above, but gives less info.  
  * ?expand=team.schedule.next Returns details of the upcoming game for a team  
  * ?expand=team.schedule.previous Same as above but for the last game played  
  * ?expand=team.stats Returns the teams stats for the season  
  * ?expand=team.roster&season=20142015 Adding the season identifier shows the roster for that season  
  * ?teamId=4,5,29 Can string team id together to get multiple teams  
  * ?stats=statsSingleSeasonPlayoffs Specify which stats to get. Not fully sure all of the values  

Examples of arguments:  

  * `ID = 20`
  * `ID = "Calgary Flames"`  
  * `expand = "person.names"`  
  * `teamId = "4, 5, 29"`  
  * `stats = "statsSingleSeasonPlayoffs"`  

Note: If you would like information for more than one team, enter `teamId`. If you want specific information, such as roster, enter `ID` and select one team at a time. For the `ID` argument, you can enter either the `teamId` or `teamName`.  

For more information about these modifiers, see [the documentation](https://gitlab.com/dword4/nhlapi/-/blob/master/stats-api.md).


```{r, message = FALSE}
baseurl_stats <- "https://statsapi.web.nhl.com/api/v1/teams"
getStats <- function(ID = "", expand = "", teamID = "", stats = ""){
  if (ID != ""){
      if (is.character(ID)) {
        ID <- index$teamId[index$teamName == ID]
      }
    baseurl_stats <- paste0("https://statsapi.web.nhl.com/api/v1/teams/", ID)
  }
  if (teamID != ""){
    fullurl <- paste0("https://statsapi.web.nhl.com/api/v1/teams?teamId=", teamID)
  }
  
  if (expand != ""){
    fullurl <- paste0(baseurl_stats, "?expand=", expand)
  } else if (stats != ""){
    fullurl <- paste0(baseurl_stats, "?stats=", stats)
  }
  stats <- GET(fullurl) %>% content("text") %>% fromJSON(flatten = TRUE)
  if (str_detect(expand, "team.roster")){
    stats_new <- stats$teams$roster.roster[[1]]
    stats <- cbind(stats$teams[,c(1,2,10)], stats_new)
  } else if (expand == "team.schedule.next"){
    if (is.null(stats$teams$nextGameSchedule.dates)){
      stop("No information is available")
    } else {
    stats_new <- stats$teams$nextGameSchedule.dates
    stats <- cbind(stats$teams[1,c(1,2,10)], stats_new)
    }
  } else if (expand == "team.schedule.previous"){
    if (is.null(stats$teams$previousGameSchedule.dates)){
      stop("No information is available")
    } else {
    stats_new <- stats$teams$previousGameSchedule.dates
    stats <- cbind(stats$teams[,c(1,2,10)], stats_new)
    }
  } else if (expand == "team.stats"){
    stats_new <- stats$teams$teamStats[[1]]$splits[[1]]
    stats <- cbind(stats$teams[1,c(1,2,11)], stats_new)
  } else {
    stats <- stats$teams
  }
  if (is.null(stats)) {
    stop("No information is available")
  } 
  return(as.data.frame(stats))
}
# getStats(ID = 20, expand = "team.stats")
# getStats(ID = "Calgary Flames", expand = "person.names")
getStats(ID = 2, expand = "team.schedule.next") %>% tbl_df()
# getStats(ID = 20, expand = "team.roster&season=20102011") %>% tbl_df()
# getStats(ID = 53, expand = "team.roster") 
getStats(teamID = "4,5,29") %>% tbl_df()
# getStats(ID = 54, stats = "statsSingleSeasonPlayoffs") %>% tbl_df()

```
### A wrapper function for all the functions above  

Endpoints:  

  * franchise (getFran)  
  * team total (getFranTeamTot)  
  * season record (getFranSeaRec)  
  * goalie record (getFranGoaRec)  
  * skater record (getFranSkaRec)  
  * stats (getStats)  
    + expand = "team.roster"  
    + expand = "person.names"  
    + expand = "team.schedule.next"  
    + expand = "team.schedule.previous"  
    + expand = "team.stats"  
    + expand = "team.roster", season = "20142015"  
    + teamId = "4, 5, 29"  
    + stats = "statsSingleSeasonPlayoffs"  

See more details in the documentation of [NHL records](https://gitlab.com/dword4/nhlapi/-/blob/master/records-api.md) and [NHL stats](https://gitlab.com/dword4/nhlapi/-/blob/master/stats-api.md).

```{r, message = FALSE}
# add a 
nhlFun <- function(endpoints, ...){
  if (endpoints == "franchise") {
    getFran()
  } else if (endpoints == "team total") {
    getFranTeamTot()
  } else if (endpoints == "season record"){
    getFranSeaRec(...)
  } else if (endpoints == "goalie record"){
    getFranGoaRec(...)
  } else if (endpoints == "skater record"){
    getFranSkaRec(...)
  } else if (endpoints == "stats"){
    getStats(...)
  } else {
    stop("Please enter a valid endpoint")
  }
}
# nhlFun(endpoints = "skater record", 20) %>% tbl_df()
nhlFun(endpoints = "stats", ID = "Calgary Flames", expand = "person.team") %>% tbl_df()
nhlFun(endpoints = "stats", ID = 2, expand = "team.schedule.next") %>% tbl_df()
```

## Exploratory Data Analysis  

### Retrieve Information

Now we demonstrate how to use functions above to do exploratory data analysis. We will first retrieve data from two endpoints: **team total** and **person.names**, remove some columns we don't need, and combine the two objects. Then we'll add two new variables: `winPercent`: the proportion of wins among all games played and `homeWinPercent`: the proportion of wins of home games among all wins. We also create a subset to include fewer columns and transform the data from wide to long. This subset shows the numbers of wins, losses, and ties in various situations for each division.  

```{r, message = FALSE}
# perhaps use the wrapper function here
franTot <- nhlFun(endpoint = "team total")
franTot <- franTot %>% select(-c("id", "activeFranchise", "firstSeasonId", "lastSeasonId"))
franStats <- nhlFun(endpoint = "stats", expand = "person.names") 
franStats <- franStats %>% select(c("locationName", "firstYearOfPlay", "franchiseId", "venue.city", "venue.timeZone.id", "venue.timeZone.tz", "division.name", "conference.name"))
# create two variables: winPercent and homeWinPercent 
combined <- inner_join(franTot, franStats, by = "franchiseId") %>% mutate(winPercent = wins / gamesPlayed, homeWinPercent = homeWins / (homeWins + homeLosses + homeOvertimeLosses + homeTies))
head(combined)

# create a subset for numbers of games lost or won
subset <- combined %>% select(starts_with("home"), starts_with("road"), "division.name", -"homeWinPercent") %>% gather(-"division.name", key = "type", value = "game") %>% group_by(division.name, type) %>% summarise(sum = sum(game, na.rm = TRUE))
```

### Summaries  

Now we have the data, we can see some summary statistics for continuous variables. For penalty minutes, it is interesting to see that, for all the variables listed, the medians are much smaller than the means, suggesting that the distributions are right skewed. Perhaps a few teams have larger than average values, pulling the means upwards. A plot on penalty minutes shows the right-skewedness.  

Two contingency tables show the counts of two categorical variables. The first contingency table shows that 1926, 1967, and 1979 added the most number of teams. The second contingency table show that the same numbers of teams play in the regular seasons and playoffs in the Atlantic, Central, and Pacific seasons. Somehow one team in the Metropolitan division did not show up in the playoffs, whether the data is missing or the team really did not make it into the playoffs.  

```{r, message = FALSE}
# summaries
apply(combined[,c(3, 6:12, 14:18)], FUN = summary, MARGIN = 2)
# contingency table
table(combined$division.name, combined$firstYearOfPlay)
table(combined$division.name, combined$gameTypeId)
```


### Visualize Data

Now we have the data and summaries, we can make some plots to visualize the data.  

In the first scatter plot, we can see that as a lot of teams play low number of games, and a few teams play a lot more number of games. The winning percentage increases slightly as the number of games played increases, but the slope is small and the difference might not be significant. The colors of dots represent different divisions, and it does not appear that there are different patterns for different divisions.The second plot looks at each division separately; again the patterns do not seem to differ.  

The third plot is similar, but the y-axis shows the winning percentage of home games, and the colors represent different types of games (regular season vs. playoffs). It does not appear that the number of games played and the winning percentage of home games are correlated, as the slope is close to zero.  

```{r, message = FALSE}
# scatter plot of homeWins and roadWins
ggplot(combined, aes(x = gamesPlayed, y = winPercent)) + geom_point(aes(color = division.name), position = "jitter") + geom_smooth(method = lm, color = "blue") + scale_color_discrete()

ggplot(combined, aes(x = gamesPlayed, y = winPercent)) + geom_point(position = "jitter") + geom_smooth(method = lm, color = "blue") + facet_wrap(~ division.name)

ggplot(combined, aes(x = gamesPlayed, y = homeWinPercent)) + geom_point(aes(color = as.factor(gameTypeId)), position = "jitter") + geom_smooth(method = lm, color = "blue") + scale_color_discrete(name = "Game Type", labels = c("regular season", "playoffs"))
```

The histogram shows that the center of winning percentages of all teams is around 0.5, and the distribution is left-skewed. Therefore most of the teams have winning percentages between 0.3 and 0.6, but there are a few teams that have much lower winning percentage.  

```{r}
# histogram of winPercent
ggplot(combined, aes(x = winPercent)) + geom_histogram(bins = 30, aes(y = ..density..)) + geom_density(kernel = "gaussian", color = "red", alpha = 0.5, fill = "green")
```

Next we look at the total numbers of games played for each division. The boxplots show that, on average, the Central division play the most games, and the Pacific division play the fewest. However, the variation is large, and therefore the differences might not be significant.  

```{r, message = FALSE}
# boxplots of gamesPlayed by division
ggplot(combined, aes(x = division.name, y = gamesPlayed)) + geom_boxplot() + geom_jitter(aes(color = venue.timeZone.tz))
```

Now we look at two barplots. The first one shows wins, losses, and ties of each division. We see that the Atlantic division play the largest number of games and the Pacific division play the smallest number of games. Overall, most results are from wins and losses, and ties and overtime occupy a small number for all divisions. The second barplots show that all divisions have higher winning percentage in playoffs than in regular season. The differences among divisions might not be significant.  

```{r, message = FALSE}
# barplot of gamePlayed
ggplot(subset, aes(y = sum, fill = type)) + geom_bar(position = "stack", stat = "identity", aes(x = division.name))

ggplot(combined, aes(x = division.name, y = winPercent, fill = as.factor(gameTypeId))) + geom_bar(position = "dodge", stat = "identity") + scale_fill_discrete(name = "Game Type", labels = c("regular season", "playoffs"))
```

Lastly, we look at a frequency plot of penalty minutes. It clearly shows that the distribution is right-skewed. 20 teams have no penalty minutes, and a few teams have penalty minutes of more than 50,000.  

```{r, message = FALSE}
# frequency plot for penalty minutes
ggplot(combined, aes(x = penaltyMinutes)) + geom_freqpoly()
```

